Asymmetric relationships and their effects on coexistence

### 2023-07-31 paper/Paper.Rnw
editted new additions explaining intuition of ATA effects, now done, moving on

%Intro the paper by intro'ing ATAs
At the core of ecology is the study of relationships between organisms and their environments.
Ecologists may ask, for instance, how local environment relates to 
population growth rate. To study such relationships, ecologists 
very commonly use metrics such as correlation and covariance. But these standard measures of the strength 
and direction of association between variables often do not tell the whole story 
\citep{Ghosh_2020_AER}.
To illustrate, the three pairs of variables in Fig. \ref{fig:pedagog1}a-c all have the same 
Pearson correlation despite striking differences in association structure: Fig. \ref{fig:pedagog1}b shows a 
symmetric type of association, while Fig. \ref{fig:pedagog1}a,c exhibit what has previously been called 
``asymmetric tail association'' [ATA; \citealp{Ghosh_2020_AER}]. 
Metrics of association other than correlations and covariance are also often used (e.g., various forms of 
regression), but correlations are common, and other common methods also reveal only a small portion 
of the information content of the relationship between variables \citep{Nelsen2006}.
ATAs and related ideas had previously been seldom considered in ecology, but recently several 
insights have been gained by studying how features of associations beyond standard measures 
impact various ecological phenomena \citep{deValpine_2014, Anderson_2019, Popovic_2019, Ghosh_2020_AER, Ghosh_2020_Eco,
Ghosh_2020_ecolevol, Ghosh_2021}. For instance, \cite{Ghosh_2020_Eco} argued that extinction risks can become elevated 
when metapopulations are subject to ATAs, e.g., if environmental ``catastrophes'' are more spatially synchronous 
across component populations
than are years of good environment, limiting the potential for rescue effects. 
And \cite{Ghosh_2021} showed that ATAs can alter the stability of 
ecosystem functioning. Fig. \ref{fig:pedagog1}d,e show contrasting examples of ATAs in nature using 
plankton population density time series; many other examples have also been previously explored 
\citep{Ghosh_2020_AER,Ghosh_2020_ecolevol,Ghosh_2021}. To our knowledge, the implications of ATAs for 
competition and coexistence have not been investigated. Doing so is the goal of this study.

%Intro to competition, coexistence, MCT, etc.
Competition, coexistence, and the maintenance of biodiversity are research topics at the core of modern ecology 
\citep{MacArthur_1958, Hutchinson_1961, Schmida_1984}. The competitive exclusion principle states 
that two species competing for the same limiting resource cannot coexist at constant densities - one will 
instead outcompete the other and drive it extinct \citep{Hardin_1960}.
Modern coexistence theory [MCT; \cite{Chesson_1994,Chesson_2000,Barabas_2018}] is a 
framework for understanding how species nevertheless coexist, and biodiversity is maintained, in real ecosystems,
through the niche partitioning effects of environmental variation.
MCT breaks down coexistence into the contributions of various mechanisms, with names such as ``storage effects'' \citep{Chesson_1994} which will be familiar to some readers (storage effects are also reviewed conceptually below and 
defined formally in Theory). MCT has been
applied to several systems, including annual \citep{Angert_2009, Hallett_2019} and perennial plants \citep{Adler_2006, Adler_2010}, 
tropical trees \citep{Usinowicz_2012}, plankton \citep{Caceres_1997, Descamps_2005, Letten2018}, and corals \citep{Alvarez-Noriega_2020}. 
%Narawani_2013
Although MCT is somewhat mathematically complex, recent extensions make it more accessible \citep{Ellner_2016,Ellner_2019}.

@@

%Lottery model and results based on it
To further demonstrate the plausibility of ATAs altering coexistence between species, we carried out 
simple simulations using a two-species model, the well-known lottery model. The simulations described in this
paragraph are sufficient to demonstrate that ATAs influence coexistence in some manner; the
precise nature of that influence is the subject of the rest of the paper.
The lottery model, which was originally introduced as a model of coral reef fish competing for a set of $N$ breeding territories, 
is as follows.
Letting $N_i(t)$ denote the adult population density of species $i=1,2$ at time $t$, and defining $N = N_1(t)+N_2(t)$,
model equations are
\begin{equation}
N_i(t+1)=(1-\delta)N_i(t)+\delta N \frac{B_i(t)N_i(t)}{B_1(t)N_1(t)+B_2(t)N_2(t)}  \label{model_eq}
\end{equation}
for $i=1,2$. Here, $\delta$ is an adult mortality rate, and $B_i(t)$ is the fecundity of species $i$ at time $t$. 
The model postulates that adults die at rate $\delta$ at each time step, and are replaced by juveniles 
in proportion to the reproductive outputs of the two species that year. Note that $N = N_1(t)+N_2(t)$ 
is constant through time. 
We assume for simplicity that the random variables $(B_1(t),B_2(t))$ are independent and identically 
distributed (iid) through time. In three distinct simulations (Fig. \ref{fig:pedagog1}f-h), 
$B_1$ and $B_2$ were, respectively, 
left- (Fig. \ref{fig:pedagog1}a), right- (Fig. \ref{fig:pedagog1}c) and symmetrically (Fig. \ref{fig:pedagog1}b)
tail associated, while the same marginal distributions for these fecundities were used in all 
simulations (see Methods
for details). Thus, the model setup was identical in the three simulations except for different ATAs. 
ATAs substantially influenced aspects of species coexistence in these simulations (Fig. \ref{fig:pedagog1}f-h).
Notation for the whole paper is summarized in Table \ref{tab:notation}.
Typesetting procedures used by publishers of ecological journals render mathematics sub-optimally,
so we have appended a latex version of this main text to the end of the 
Supporting Information (SI). Readers interested in the mathematical content
of the study should read that version.
%***DAN: Don't forget to do this

%Some intuition
We next give some initial intuition about how ATA effects on coexistence work, 
providing additional intuition in the Discussion.
It is known that storage effects can occur if a buffering mechanism reduces population 
losses in bad years so that
gains in good years can more than make up for them. 
Bad years happen when the negative effects of $C$ outweigh any 
positive effects of $E$; but this can occur
both 1) when $E$ is moderately favorable but $C$ is very strong, as well as 2) when $E$ is 
unfavorable and $C$ is not too weak.
Conversely, good years happen when positive effects of $E$ outweigh negative effects of $C$;
but this can occur 1) when $E$ is favorable and $C$ is not too strong,
as well as 2) when $E$ is moderately unfavorable but $C$ is very weak. 
It turns out that the potential for storage events to occur when 
$E$ and $C$ are both higher than average (the cases 1, above); versus the potential  
when $E$ and $C$ are
both lower than average (the cases 2) are not necessarily equal -- storage potential may differ
across the spectrum from 1 to 2.
ATA effects quantify the extent to which asymmetries of association between $E$ and 
$C$ are aligned with differences in the potential for storage across the spectrum from 1 to 2.
If $E$ and $C$ are largely unassociated in the parts of their distribution where
storage potential is high, ATA effects will be high, and vice versa.

@@

%Para continuing to try to build intuition about how ATAs work
We return to the task, begun in the Introduction, of building intuition for how
ATA effects work. 
Just as for storage effects [\cite{Chesson_1994}, Box 2 of \cite{StumpVasseur2023}], 
temporal niche differentiation is necessary for ATA
effects (SI section \ref{SI-SIsect:temporal_niche_diff}). 
If temporal niche differentiation occurs
but is partial, as will be typical, $E_i$ and $C_{i \bs i}$ for a 
species $i$ will be positively but imperfectly associated, 
and the association may be asymmetric. 
Storage events occur for species $i$ when the positive effects of $E_i$ overwhelm the negative 
effects of $C_{i \bs i}$, producing large values of $r_i(E_i,C_{i \bs i})$
which are not completely offset by declines at other times. 
For instance, for the lottery model,
$r_i(E_i,C_{i \bs i})=\log(1-\delta+E_i/C_{i \bs i})$ can get large whenever
$C_{i \bs i}$ is a small fraction of $E_i$, but can never get smaller
than $\log(1-\delta)$. 
The potential for storage events can differ when $E_i$ and $C_{i \bs i}$ are both
above their median values compared to when they are both below median. For instance,
the ratio $E_i/C_{i \bs i}$, and therefore the growth rate 
$\log(1-\delta+E_i/C_{i \bs i})$, can both get very big, corresponding to
rapid growth, if $C_{i \bs i}$ is close to $0$, even if $E_i$ is fairly small.
In contrast, if $E_i$ and $C_{i \bs i}$ are both in the right portions of their distributions, 
so that $C_{i \bs i}$ is bounded away from $0$, there may be much sharper limits on the 
ratio $E_i/C_{i \bs i}$ and the corresponding growth rate.
ATA effects quantify the extent to which asymmetries of association between $E$ and 
$C$ are aligned with varying potential for storage across the $E,C$ distribution. 
ATA effects help quantify different degrees of temporal niche differentiation 
occurring under different sets of environmental conditions, possibly via
different mechanisms.
Details and examples of this intuition 
are in SI section \ref{SI-SIsect:intuition_examples} and SI 
Figs \ref{SI-fig:beta_fecundidities_explanatory}-\ref{SI-fig:lognormal_fecundidities_explanatory}, 
where comparisons are also made between our two lottery models.

>>>>

%Intro the paper by intro'ing ATAs
At the core of ecology is the study of relationships between organisms and their environments.
Ecologists may ask, for instance, how local environment relates to 
population growth rate. To study such relationships, ecologists 
very commonly use metrics such as correlation and covariance. But these standard measures of the strength 
and direction of association between variables often do not tell the whole story 
\citep{Ghosh_2020_AER}.
To illustrate, the three pairs of variables in Fig. \ref{fig:pedagog1}a-c all have the same 
Pearson correlation despite striking differences in association structure: Fig. \ref{fig:pedagog1}b shows a 
symmetric type of association, while Fig. \ref{fig:pedagog1}a,c exhibit what has previously been called 
``asymmetric tail association'' [ATA; \citealp{Ghosh_2020_AER}]. 
Metrics of association other than correlations and covariance are also often used (e.g., various forms of 
regression), but correlations are common, and other common methods also reveal only a small portion 
of the information content of the relationship between variables \citep{Nelsen2006}.
ATAs and related ideas had previously been seldom considered in ecology, but recently several 
insights have been gained by studying how features of associations beyond standard measures 
impact various ecological phenomena \citep{deValpine_2014, Anderson_2019, Popovic_2019, Ghosh_2020_AER, Ghosh_2020_Eco,
Ghosh_2020_ecolevol, Ghosh_2021}. For instance, \cite{Ghosh_2020_Eco} argued that extinction risks can become elevated 
when metapopulations are subject to ATAs, e.g., if environmental ``catastrophes'' are more spatially synchronous 
across component populations
than are years of good environment, limiting the potential for rescue effects. 
And \cite{Ghosh_2021} showed that ATAs can alter the stability of 
ecosystem functioning. Fig. \ref{fig:pedagog1}d,e show contrasting examples of ATAs in nature using 
plankton population density time series; many other ecological examples of ATAs have also been explored 
\citep{Ghosh_2020_AER,Ghosh_2020_ecolevol,Ghosh_2021}. To our knowledge, the implications of ATAs for 
competition and coexistence have not been investigated. Doing so is the goal of this study.

%Intro to competition, coexistence, MCT, etc.
Competition, coexistence, and the maintenance of biodiversity are research topics at the core of modern ecology 
\citep{MacArthur_1958, Hutchinson_1961, Schmida_1984}. The competitive exclusion principle states 
that two species competing for the same limiting resource cannot coexist at constant densities - one will 
instead outcompete the other and drive it extinct \citep{Hardin_1960}.
Modern coexistence theory [MCT; \cite{Chesson_1994,Chesson_2000,Barabas_2018}] is a 
framework for understanding how species nevertheless coexist, and biodiversity is maintained, in real ecosystems,
through the niche partitioning effects of environmental variation.
MCT breaks down coexistence into the contributions of various mechanisms, with names such as ``storage effects'' \citep{Chesson_1994}; storage effects are reviewed conceptually below and will be
defined formally in the Theory section. MCT has been
applied to several systems, including annual \citep{Angert_2009, Hallett_2019} and perennial plants \citep{Adler_2006, Adler_2010}, 
tropical trees \citep{Usinowicz_2012}, plankton \citep{Caceres_1997, Descamps_2005, Letten2018}, and corals \citep{Alvarez-Noriega_2020}. 
%Narawani_2013
Although MCT is somewhat mathematically complex, recent extensions make it more accessible \citep{Ellner_2016,Ellner_2019}.

@@

%Lottery model and results based on it
To further demonstrate the plausibility of ATAs altering coexistence between species, we carried out 
simple simulations using a two-species model, the well-known lottery model. The simulations described in this
paragraph are sufficient to demonstrate that ATAs influence coexistence in some manner; the
precise nature of that influence is the subject of the rest of the paper.
The lottery model, which was originally introduced as a model of coral reef fish competing for a set of $N$ breeding territories, 
is as follows.
Letting $N_i(t)$ denote the adult population density of species $i=1,2$ at time $t$, and defining $N = N_1(t)+N_2(t)$,
model equations are
\begin{equation}
N_i(t+1)=(1-\delta)N_i(t)+\delta N \frac{B_i(t)N_i(t)}{B_1(t)N_1(t)+B_2(t)N_2(t)}  \label{model_eq}
\end{equation}
for $i=1,2$. Here, $\delta$ is an adult mortality rate, and $B_i(t)$ is the fecundity of species $i$ at time $t$. 
The model postulates that adults die at rate $\delta$ at each time step, and are replaced by juveniles 
in proportion to the reproductive outputs of the two species that year. Note that $N = N_1(t)+N_2(t)$ 
is constant through time. 
We assume for simplicity that the random variables $(B_1(t),B_2(t))$ are independent and identically 
distributed (iid) through time. In three distinct simulations (Fig. \ref{fig:pedagog1}f-h), 
$B_1$ and $B_2$ were, respectively, 
left- (Fig. \ref{fig:pedagog1}a), right- (Fig. \ref{fig:pedagog1}c) and symmetrically (Fig. \ref{fig:pedagog1}b)
tail associated, while the same marginal distributions for these fecundities were used in all 
simulations (see Methods
for details). Thus, the model setup was identical in the three simulations except for different ATAs. 
ATAs substantially influenced aspects of species coexistence in these simulations (Fig. \ref{fig:pedagog1}f-h).
Notation for the whole paper is summarized in Table \ref{tab:notation}.
Typesetting of mathematics by ecological journals is sub-optimal,
so we appended a latex version of this main text to the end of the 
Supporting Information (SI). Readers interested in the mathematical aspects should read that version.
%***DAN: Don't forget to do this

%Some intuition
We next give some initial intuition about how ATA effects on coexistence work, 
providing additional intuition in the Discussion.
It is known that storage effects can occur if a buffering mechanism reduces population 
losses in bad years so that
gains in good years can more than make up for them. 
Bad years happen when the negative effects of $C$ outweigh any 
positive effects of $E$; but this can occur
both 1) when $E$ is moderately favorable but $C$ is very strong, as well as 2) when $E$ is 
unfavorable and $C$ is not too weak.
Conversely, good years happen when positive effects of $E$ outweigh negative effects of $C$;
but this can occur 1) when $E$ is favorable and $C$ is not too strong,
as well as 2) when $E$ is moderately unfavorable but $C$ is very weak. 
It turns out that the potential for storage events to occur when 
$E$ and $C$ are both higher than average (the cases 1, above); versus the potential  
when $E$ and $C$ are
both lower than average (the cases 2) are not necessarily equal -- storage potential may differ
across the spectrum from 1 to 2.
ATA effects quantify the extent to which asymmetries of association between $E$ and 
$C$ are aligned with differences in the potential for storage across the spectrum from 1 to 2.
If $E$ and $C$ are largely unassociated in the parts of their distribution where
storage potential is high, ATA effects will be high. ATA effects quantify how temporal niche partitioning
varies across $EC$ spectrum spanned by the cases 1-2 sketched above.

@@

%Para continuing to try to build intuition about how ATAs work
We return to the task, begun in the Introduction, of building intuition for how
ATA effects work. 
Just as for storage effects [\cite{Chesson_1994}, Box 2 of \cite{StumpVasseur2023}], 
temporal niche differentiation is necessary for ATA
effects (SI section \ref{SI-SIsect:temporal_niche_diff}). 
If temporal niche differentiation occurs
but is partial, as will be typical, $E_i$ and $C_{i \bs i}$ for a 
species $i$ will be positively but imperfectly associated, 
and the association may be asymmetric. 
Storage events occur for species $i$ when the positive effects of $E_i$ overwhelm the negative 
effects of $C_{i \bs i}$, producing large values of $r_i(E_i,C_{i \bs i})$
which are not completely offset by population declines at other times. 
For instance, for the lottery model,
$r_i(E_i,C_{i \bs i})=\log(1-\delta+E_i/C_{i \bs i})$ can get large whenever
$C_{i \bs i}$ is a small fraction of $E_i$, but can never get smaller
than $\log(1-\delta)$. 
The potential for storage events can differ when $E_i$ and $C_{i \bs i}$ are both
above their median values compared to when they are both below median. For instance,
the ratio $E_i/C_{i \bs i}$, and therefore the growth rate 
$\log(1-\delta+E_i/C_{i \bs i})$, can both get very big, corresponding to
rapid growth, if $C_{i \bs i}$ is close to $0$, even if $E_i$ is fairly small.
In contrast, if $E_i$ and $C_{i \bs i}$ are both in the right portions of their distributions, 
so that $C_{i \bs i}$ is bounded away from $0$, there may be much sharper limits on the 
ratio $E_i/C_{i \bs i}$ and the corresponding growth rate.
ATA effects quantify the extent to which asymmetries of association between $E$ and 
$C$ are aligned with varying potential for storage across the $E,C$ distribution. 
ATA effects help quantify different degrees of temporal niche differentiation 
occurring under different sets of environmental conditions, possibly via
different mechanisms.
Further details and examples of this intuition 
are in SI section \ref{SI-SIsect:intuition_examples} and SI 
Figs \ref{SI-fig:beta_fecundidities_explanatory}-\ref{SI-fig:lognormal_fecundidities_explanatory}, 
where comparisons are also made between our two lottery models.
